sfislands: An R Package for Accommodating Islands and Disjoint Zones in Areal Spatial Modelling

### 2024-03-15 article.tex
revisions to article

For R users, the \CRANpkg{spdep} package (Bivand et al. 2005) has long been popular for the
creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}
structures and their compatibility with the \CRANpkg{tidyverse} (Wickham et al. 2019), the \CRANpkg{sfdep} (Pebesma 2018) package (Parry and Locke 2024) has presented the same
functionality in addition to extra features in a tidy structure based on
lists in a similar way to \CRANpkg{sf}.

@@

The most commonly used neighbourhood structure is one based on first-order queen contiguity, where units are considered neighbours if they
share at least a vortex of boundary. However, as the name suggests, this
will lead to problems when non-contiguous units such as islands or
exclaves are present. Less obviously, depending on how the geographic
units are described, areas on either sides of rivers may be
inappropriately classified as neighbours or not neighbours. Furthermore,
the presence of infrastructure such as tunnels, bridges or ferry
services might be satisfactory to meet our hypothesis of the required
degree of connectivity to be considered neighbours.

The aim of \texttt{sfislands} is to deal with the situations described above in a
convenient and open manner. It allows us to set up a structure, quickly
map it, and then examine it. The structure can then be edited and the process re-iterated until we have described a
spatial relationship structure with which we are satisfied.

The above can be considered as the \emph{pre-functions} of the package. A
further set of feature of \texttt{sfislands}, which we refer to as \emph{post-functions},
are for use with the \CRANpkg{mgcv} package (Wood 2011). It is straightforward to use
\texttt{sfislands} neighbourhood structures in models with \CRANpkg{mgcv} but it can be quite awkward
to extract the estimates. These \emph{post-functions} conveniently provide the estimates in tidy format of random effects and intrinsic conditional autoregressive (ICAR) components from such models and give the ability to quickly map them for visualisation purposes.

@@

In this paper, we will look at three examples to show different use-cases for
\texttt{sfislands}. The first example focuses on earthquakes in
Indonesia. It shows a scenario where all of the functions are used, from
setting up contiguities to modelling and examining the estimates of the
model.

@@

It greatly speeds up the workflow of fitting areal spatial models.

\begin{itemize}
\item
  \textbf{pre-functions} for setting up contiguities:

  \begin{itemize}
  \item
    it addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data. \texttt{sfislands} will include a contiguity for all units,
  \item
    it gives tools to immediately visualise this structure as a map,
  \item
    these maps are created using \CRANpkg{ggplot2} (Wickham 2016), which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax,
  \item
    as the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence,
  \item
    connections made by the package which are not contiguities can be accessed to ensure openness in the process.
  \end{itemize}
\item
  For modelling:

  \begin{itemize}
  \tightlist
  \item
    these neighbourhood structures can be used in modelling packages such as \CRANpkg{mgcv}, \CRANpkg{brms}. (Bürkner 2017), \pkg{r-inla} (Bakka et al. 2018) and more.
  \end{itemize}
\item
  \textbf{post-functions} for \texttt{mgcv} models:

  \begin{itemize}
  \item
    it simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models),
  \item
    these effects can be quickly visualised in \CRANpkg{ggplot2} maps.
  \end{itemize}
\end{itemize}

@@

The first group of functions, shown in Table \ref{tab:prefunc-latex}, deals with the creation
of a contiguity structure in the presence of discontiguities. The
resultant structure can be quickly mapped to check if it is
satisfactory. Connections can be manually added or removed by name or
index number. By an iterative process of changes and examination of a
quick map, a satisfactory structure can be decided upon.

Table \ref{tab:prefunc-latex} shows these functions and their purpose.

@@

\caption{\label{tab:prefunc-latex}Pre-functions: setting up a neighbourhood structure.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l|>{\raggedright\arraybackslash}p{7cm}}
\hline
function & purpose\\
\hline
st\_bridges() & create a neighbourhood contiguity structure, with a k-nearest neighbours condition for islands\\
\hline
st\_quickmap\_nb() & check structure visually on map\\
\hline
st\_check\_islands() & check the contiguities which have been assigned to islands\\
\hline
st\_manual\_join\_nb() & make manual changes by adding connections\\
\hline
st\_manual\_cut\_nb() & make manual changes by removing connections\\
\hline
\end{tabular}
\end{table}

They deal with the creation
of a contiguity structure in the presence of discontiguities. The
resultant structure can be quickly mapped to check if it is
satisfactory. Connections can be manually added or removed by name or
index number. By an iterative process of changes and examination of a
quick map, a satisfactory structure can be decided upon.
Going through each function in more detail:

Use the rectangles shown in Figure \ref{fig:rects1} as a simple demonstration. Rectangles 1-3
are contiguous while 4 and 5 are islands.

@@

This function requires as arguments an \CRANpkg{sf} dataframe and the name of a column of unique identifiers, preferably names, of each spatial unit. The k-nearest neighbours for connecting islands can then be specified. The output is a \emph{named} neighbourhood structure in either list or matrix form as desired, which can be either a standalone object or included as an additional column in the original \CRANpkg{sf} dataframe. While it is not necessary in all modelling packages for the neighbourhood list or matrix to be \emph{named}, it is good practice to do so and is mandatory when using, for example, \CRANpkg{mgcv}.

In the following examples, we have chosen to ignore islands with the argument \texttt{remove\_islands\ =\ TRUE} and to return a list and matrix structure respectively by specifying this in the \texttt{nb\_structure} argument and choosing \texttt{add\_to\_dataframe\ =\ FALSE}:

@@

In these examples, we choose to join islands to their 1 nearest
neighbour, which is the default setting, and to return the output as a column called ``nb'' in the original
\CRANpkg{sf} dataframe (this is the default setting):

@@

\begin{verbatim}
#> Simple feature collection with 5 features and 2 fields
#> Geometry type: POLYGON
#> Dimension:     XY
#> Bounding box:  xmin: 0 ymin: 0 xmax: 6 ymax: 4
#> CRS:           NA
#>                         geometry  name      nb
#> 1 POLYGON ((0 0, 0 2, 2 2, 2 ... Rect1    2, 3
#> 2 POLYGON ((2 0, 2 2, 4 2, 4 ... Rect2 1, 3, 4
#> 3 POLYGON ((2 2, 2 4, 4 4, 4 ... Rect3 1, 2, 5
#> 4 POLYGON ((5 0, 5 1, 6 1, 6 ... Rect4       2
#> 5 POLYGON ((0.8 3, 0.8 4, 1.8... Rect5       3
\end{verbatim}

@@

\begin{verbatim}
#> Simple feature collection with 5 features and 2 fields
#> Geometry type: POLYGON
#> Dimension:     XY
#> Bounding box:  xmin: 0 ymin: 0 xmax: 6 ymax: 4
#> CRS:           NA
#>                         geometry  name nb.1 nb.2 nb.3 nb.4 nb.5
#> 1 POLYGON ((0 0, 0 2, 2 2, 2 ... Rect1    0    1    1    0    0
#> 2 POLYGON ((2 0, 2 2, 4 2, 4 ... Rect2    1    0    1    1    0
#> 3 POLYGON ((2 2, 2 4, 4 4, 4 ... Rect3    1    1    0    0    1
#> 4 POLYGON ((5 0, 5 1, 6 1, 6 ... Rect4    0    1    0    0    0
#> 5 POLYGON ((0.8 3, 0.8 4, 1.8... Rect5    0    0    1    0    0
\end{verbatim}

These structures can serve as the input to models in \CRANpkg{brms},
\pkg{r-inla}, \pkg{rstan} (Stan Development Team 2020) or \CRANpkg{mgcv}. Rather than having a separate neighbours
column, it is included as a named list or matrix in the original \CRANpkg{sf}
dataframe, in the spirit of the \CRANpkg{sfdep} package.

@@

This function will tell us openly what connections have been made which
are not based on contiguity. It gives both the name and index number of each pair of added connections.

\begin{verbatim}
st_bridges(rectangles, 
           "name", 
           link_islands_k = 1) |> 
  st_check_islands()
\end{verbatim}

@@

\begin{verbatim}
st_bridges(rectangles, "name", 
           link_islands_k = 1) |> 
  st_manual_join_nb(3,4) |> 
  st_quickmap_nb(nodes = "numeric")
\end{verbatim}

@@

\begin{verbatim}
st_bridges(rectangles, "name", 
           link_islands_k = 1) |> 
  st_manual_join_nb(3,4) |> 
  st_manual_cut_nb("Rect1","Rect2") |> 
  st_quickmap_nb(nodes = "numeric")
\end{verbatim}

@@

The use of these structures is particularly common in CAR-type models.
These are often implemented in a Bayesian framework using
\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. The pre-functions will output the
neighbourhood structure in the desired format for use in any of these
frameworks. A convenient frequentist alternative is to use the
\CRANpkg{mgcv} package. It has the functionality to create such models
using \texttt{bs="mrf"}. It also has the ability to combine these with a
hierarchical structure using \texttt{bs="re"}. While the outputs from the
Bayesian structures mentioned above can be extracted in the same way as
any other component of the model, it can be a bit awkward to get the
estimates from \CRANpkg{mgcv} model. \texttt{sfislands} has two post-functions to conveniently extract and visualise these.

@@

\caption{\label{tab:postfunc-latex}Post-functions: tidy estimates from mgcv.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l|l}
\hline
function & purpose\\
\hline
st\_augment() & augment the original dataframe with model predictions\\
\hline
st\_quickmap\_preds() & generate quick maps of these predictions\\
\hline
\end{tabular}
\end{table}

@@

This function augments the original dataframe with the estimated means
and standard errors from the \CRANpkg{mgcv} model in a similar manner to
how the \CRANpkg{broom} (Robinson, Hayes, and Couch 2023) package operates. The \texttt{geometry} column, as per convention, remains as the last column of the augmented dataframe, while the new random effects columns are positioned immediately before it. The estimates which \texttt{st\_augment()} will extract are random effects (which are called in \CRANpkg{mgcv} with \texttt{bs=\textquotesingle{}re\textquotesingle{}}) and ICAR components (\texttt{bs=\textquotesingle{}mrf\textquotesingle{}}).

Consider the model structure described in the code below using \CRANpkg{mgcv} syntax. In this model \(y\) is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a random intercept and slope for the covariate at a \emph{region} level, and an ICAR varying intercept and slope at a lower \emph{sub-region} level.

\begin{verbatim}
mgcv::gam(
  y ~ covariate +                   # fixed intercept and effect for covariate  
    s(region, bs = "re") +                  # random intercept at level region
    s(region, covariate, bs = "re") +          # random slopes at level region
    s(sub-region, 
      bs = 'mrf',
      xt = list(nb = data$nb), 
      k = k) +                    # ICAR varying intercept at level sub-region
    s(sub-region, by = covariate, 
      bs = 'mrf',
      xt = list(nb = data$nb), 
      k = k),           # ICAR varying slope for covariate at level sub-region
  data = data, 
  method = "REML")
\end{verbatim}

When labelling the new columns which are augmented to the original dataframe from such a model, \texttt{st\_augment()} follows the formula syntax of the \CRANpkg{lme4} (Bates et al. 2015) package, where the \emph{pipe symbol} (\texttt{\textbar{}}) indicates ``\emph{grouped by}''. Table \ref{tab:staugtab-latex} shows how the augmented columns in this scenario would be labelled. Each column name begins with either \texttt{random.effect.} or \texttt{mrf.smooth.} as appropriate. An additional column is also added for the standard error of each estimate. These columns are named as above but with \texttt{se.} prepended (e.g.~\texttt{se.random.effect.region}).

@@

\caption{\label{tab:staugtab-latex}The naming procedure for augmented columns from different mgcv structures.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l|l}
\hline
mgcv syntax & column name\\
\hline
s(region, bs = 're') & random.effect.region\\
\hline
s(region, covariate, bs = 're') & random.effect.covariate|region\\
\hline
s(sub-region, bs = 'mrf', xt = list(nb = data\$nb) & mrf.smooth.sub-region\\
\hline
s(sub-region, by = covariate, bs = 'mrf', xt = list(nb = data\$nb) & mrf.smooth.covariate|sub-region\\
\hline
\end{tabular}
\end{table}

@@

These estimates can then be quickly mapped. As it is possible to include more than 1 varying component, the output of this function is a list of plots. They can be viewed individually by indexing, or all at one using the \texttt{plotlist=} argument from the \texttt{ggarrange()} function which is part of the \CRANpkg{ggpubr} (Kassambara 2023) package. We will see examples of this in the following examples. The maps are automatically titled and subtitled according to the type of effect. For example, the map showing \texttt{random.effect.region} will have ``\emph{region}'' as its title and ``\emph{random.effect}'' as its subtitle.

@@

The data for this section have been taken from the \href{https://earthquake.usgs.gov/earthquakes/search/}{USGS earthquake
catalogue API}. The
datasets and an explanation are available at
\url{https://github.com/horankev/quake_data} and record all earthquakes in
and close to Indonesia from 1985-2023. Below is a map of Indonesia, with
other neighbouring or bordering countries filled in grey. The many local
faults which lie within 300km of the shore are shown as green lines
in Figure \ref{fig:fault-buffers}.

@@

To get an interpretable measure of the concentration of faults in any area
they are transformed from linestrings to polygons by setting a buffer of
10km around them, as shown below. Now both our faults and the sizes of provinces are in units of kilometres
squared. This means we can generate a unitless metric of what proportion of any administrative unit
is covered by these buffered faults. This measure across provinces is shown in Figure
\ref{fig:fault-conc}.

@@

Earthquake activity will be seen as the total number of earthquakes with
their epicentre in each province per unit area of that province. We have
restricted counts to earthquakes of magnitude \textgreater5.5 which is the point
at which they are often labelled as damaging.

The occurrences of these earthquakes are shown in Figure \ref{fig:quake-occur}, their total per province in Figure \ref{fig:quake-totals}, and finally, their count per square kilometre can be seen in Figure \ref{fig:quake-conc}.

@@

\caption{Earthquakes in Indonesia, 1985-2023. Categorised as medium, large or extra-large. }\label{fig:quake-occur}
\end{figure}

@@

As this is count data, we will model it as a poisson distribution with
\(\lambda\) as the mean count per province. For \(i = 1,...,n\) provinces,
the dependent variable in this model is
\begin{equation}
y_i = \text{earthquake incidence}_i = \frac{\text{earthquake count in province}_i} {\text{province area}_i}, 
\label{eq:eq1}
\end{equation}
while the explanatory variable is
\begin{equation}
x_i = \text{fault intensity}_i = \frac{\text{area of buffered faults in province}_i} {\text{province area}_i}.
\label{eq:eq2}
\end{equation}
Firstly, when excluding the incidence and just modelling counts, where
\(y_i = \text{earthquake count in province}_i\), the poisson model is of
the following form:
\begin{equation}
y_i | \lambda_i \sim \text{Pois}(\lambda_i)
\label{eq:eq3}
\end{equation}
with
\begin{equation}
E(y_i | \lambda_i) = \lambda_i.
\label{eq:eq4}
\end{equation}
We model
\begin{equation}
log(\lambda_i) = \beta_0 + \beta_1x_i + \gamma_i.
\label{eq:eq5}
\end{equation}
where \(\gamma_i\)
is a term with a correlation structure reflecting a province's location
relative to other provinces.

We can describe these relationships by setting up a neighbourhood
structure based on queen contiguity where a pair of provinces are
considered neighbours if they share at least one point of boundary. This can be modelled as a Markov random field to generate an ICAR model with a spatially varying term. Each of these terms will be correlated the others according to the neighbourhood structure we have defined.

@@

If we include an offset term (here, area) because we are more interested
in the incidence than in the actual count, then
\begin{equation}
log(\frac{\lambda_i} {\text{area}_i}) = \beta_0 + \beta_1x_i + \gamma_i
\label{eq:eq6}
\end{equation}
which is equivalent to
\begin{equation}
log(\lambda_i) = \beta_0 + \beta_1x_i + \gamma_i + log(\text{area}_i).
\label{eq:eq7}
\end{equation}
We are still modelling \(log(\lambda)\) rather than the incidence, but we
are adding an offset to adjust for differing areas. Modelling \(log(\lambda)\) and adding an offset is equivalent to
modelling incidence, and coefficients can be interpreted that way.

@@

@@

Having made some manual adjustments to the neighbourhood structure, the
new structure can be seen in Figure \ref{fig:ind-contig3}. Edge effects
have also been mitigated by additionally connecting the two extreme
provinces (1 and 23), which would otherwise have only one neighbour, to
their two next closest neighbours.

@@

We now create the ICAR model using \CRANpkg{mgcv}. We will be able to simply pass
the contiguity structure we have created as both the data and a column
of it as the neighbourhood structure.

@@

\hypertarget{workflow}{%
\subsection{Workflow}\label{workflow}}

>>>>

For R users, the \CRANpkg{spdep} package (Bivand et al. 2005) has long been popular for the
creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}
structures and their compatibility with the \CRANpkg{tidyverse} (Wickham et al. 2019), the \CRANpkg{sfdep} package (Parry and Locke 2024) has presented the same
functionality in addition to extra features in a tidy structure based on
lists in a similar way to \CRANpkg{sf}.

@@

The most commonly used neighbourhood structure is one based on first-order queen contiguity, where units are considered neighbours if they
share at least a vortex of boundary. However, as the name suggests, this
will lead to problems when non-contiguous units such as islands or
exclaves are present. Less obviously, depending on how the geographic
units are described, areas on either sides of rivers may be
inappropriately classified as neighbours or not neighbours. Furthermore,
the presence of infrastructure such as tunnels, bridges or ferry
services might be satisfactory to meet our hypothesis of the required
degree of connectivity to be considered neighbours. Again, such information will not be found in a set of polygons.

The aim of \texttt{sfislands} is to deal with the situations described above in a
convenient and open manner. It allows us to set up a structure, quickly
map it, and then examine whether or not it is consistent with our hypothesis of relationships between units. The structure can then be edited and the process re-iterated until we have described a
spatial relationship structure with which we are satisfied.

The above can be considered as the \emph{pre-functions} of the package. A
second category of features, which we refer to as \emph{post-functions},
are for use with the \CRANpkg{mgcv} package (Wood 2011). Having fit a model with \CRANpkg{mgcv}, the process of extracting estimates for certain types of effects can be somewhat awkward. These \emph{post-functions} augment the original dataframe with these estimates and their standard errors in tidy format. They also allow for quick visualisation of the output in map form.

@@

In this paper, we will look at three examples to show different use-cases for
\texttt{sfislands}. The first example focuses on earthquakes in
Indonesia. It shows a scenario where all of the functions are used, from
setting up contiguities, to modelling and examining the estimates of the
model.

@@

It greatly speeds up each stage of a proposed workflow for fitting areal spatial models.

\hypertarget{step-1-pre-functions-for-setting-up-neighbourhood-structure}{%
\subsubsection{\texorpdfstring{Step 1: \emph{Pre-functions} for setting up neighbourhood structure}{Step 1: Pre-functions for setting up neighbourhood structure}}\label{step-1-pre-functions-for-setting-up-neighbourhood-structure}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  It addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data. \texttt{sfislands} will include a contiguity for all units.
\item
  It gives tools to immediately visualise this structure as a map.
\item
  These maps are created using \CRANpkg{ggplot2} (Wickham 2016), which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax.
\item
  As the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence.
\item
  Connections which have been induced by a function from the package but are not contiguities can be accessed to ensure openness in the process.
\end{enumerate}

\hypertarget{step-2-modelling}{%
\subsubsection{Step 2: Modelling}\label{step-2-modelling}}

These neighbourhood structures can be used in modelling packages such as \CRANpkg{mgcv}, \CRANpkg{brms} (Bürkner 2017), \pkg{r-inla} (Bakka et al. 2018) and more.

\hypertarget{step-3-post-functions-for-mgcv-models}{%
\subsubsection{\texorpdfstring{Step 3: \emph{Post-functions} for \texttt{mgcv} models}{Step 3: Post-functions for mgcv models}}\label{step-3-post-functions-for-mgcv-models}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  It simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models).
\item
  These effects can be quickly visualised as \CRANpkg{ggplot2} maps.
\end{enumerate}

@@

The first group of functions, shown in Table \ref{tab:prefunc-latex}, deals with the creation
of a neighbourhood structure in the presence of discontiguities. The
resultant structure can be quickly mapped to check if it is
satisfactory. Connections can be manually added or removed by name or
index number. By an iterative process of changes and examination of a
quickly-generated guide map, a satisfactory structure can be decided upon.

@@

\caption{\label{tab:prefunc-latex}Pre-functions: setting up a neighbourhood structure.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l|>{\raggedright\arraybackslash}p{7cm}}
\hline
\textbf{function} & \textbf{purpose}\\
\hline
st\_bridges() & create a neighbourhood contiguity structure, with a k-nearest neighbours condition for islands\\
\hline
st\_quickmap\_nb() & check structure visually on map\\
\hline
st\_check\_islands() & check the contiguities which have been assigned to islands\\
\hline
st\_manual\_join\_nb() & make manual changes by adding connections\\
\hline
st\_manual\_cut\_nb() & make manual changes by removing connections\\
\hline
\end{tabular}
\end{table}

We will now go through each function in more detail using the set of rectangles shown in Figure \ref{fig:rects1} for demonstration purposes. Rectangles 1, 2 and 3 are contiguous while 4 and 5 can be viewed as ``islands''.

@@

This function requires at least two arguments: an \CRANpkg{sf} dataframe and, from that, the name of one column of unique identifiers, ideally names, of each spatial unit. It creates a neighbourhood structure where non-island units are joined by first-order queen contiguity, while island units are joined to their k-nearest neighbours. The output is a \emph{named} neighbourhood structure in either list or matrix form as desired, which can be either a standalone object or included as an additional column in the original \CRANpkg{sf} dataframe. While it is not necessary in all modelling packages for the neighbourhood list or matrix to be \emph{named}, it is good practice to do so and is mandatory when using, for example, \CRANpkg{mgcv}.

In the first two examples of using \texttt{st\_bridges()}, we have chosen to ignore islands with the argument \texttt{remove\_islands\ =\ TRUE} and to return a list and matrix structure respectively by specifying this in the \texttt{nb\_structure} argument and choosing \texttt{add\_to\_dataframe\ =\ FALSE}:

@@

Alternatively, in the following examples, we choose to join islands to their 1 nearest
neighbour, which is the default setting, and to return the output as a column called ``nb'' in the original
\CRANpkg{sf} dataframe (\texttt{add\_to\_dataframe\ =\ "TRUE"} is the default setting):

@@

\begin{verbatim}
#> Simple feature collection with 5 features and 2 fields
#> Geometry type: POLYGON
#> Dimension:     XY
#> Bounding box:  xmin: 0 ymin: 0 xmax: 6 ymax: 4
#> CRS:           NA
#>    name      nb                       geometry
#> 1 Rect1    2, 3 POLYGON ((0 0, 0 2, 2 2, 2 ...
#> 2 Rect2 1, 3, 4 POLYGON ((2 0, 2 2, 4 2, 4 ...
#> 3 Rect3 1, 2, 5 POLYGON ((2 2, 2 4, 4 4, 4 ...
#> 4 Rect4       2 POLYGON ((5 0, 5 1, 6 1, 6 ...
#> 5 Rect5       3 POLYGON ((0.8 3, 0.8 4, 1.8...
\end{verbatim}

@@

\begin{verbatim}
#> Simple feature collection with 5 features and 2 fields
#> Geometry type: POLYGON
#> Dimension:     XY
#> Bounding box:  xmin: 0 ymin: 0 xmax: 6 ymax: 4
#> CRS:           NA
#>    name nb.1 nb.2 nb.3 nb.4 nb.5                       geometry
#> 1 Rect1    0    1    1    0    0 POLYGON ((0 0, 0 2, 2 2, 2 ...
#> 2 Rect2    1    0    1    1    0 POLYGON ((2 0, 2 2, 4 2, 4 ...
#> 3 Rect3    1    1    0    0    1 POLYGON ((2 2, 2 4, 4 4, 4 ...
#> 4 Rect4    0    1    0    0    0 POLYGON ((5 0, 5 1, 6 1, 6 ...
#> 5 Rect5    0    0    1    0    0 POLYGON ((0.8 3, 0.8 4, 1.8...
\end{verbatim}

These structures can serve as the input to models in \CRANpkg{brms},
\pkg{r-inla}, \pkg{rstan} (Stan Development Team 2020) or \CRANpkg{mgcv}. \CRANpkg{brms} requires a matrix structure while \CRANpkg{mgcv} models use a list. Rather than having a separate neighbours
object, it is included in the original \CRANpkg{sf} dataframe as a named list or matrix, in the spirit of the \CRANpkg{sfdep} package.

@@

This function will show us transparently what connections have been made which
are not based on contiguity. It gives both the name and index number of each pair of added connections. In this example, two pairs have been added.

\begin{verbatim}
# show summary of non-contiguous connections in a dataframe

st_bridges(rectangles, 
           "name", 
           link_islands_k = 1) |> 
  st_check_islands()
\end{verbatim}

@@

\begin{verbatim}
# add an extra connection using numeric index

st_bridges(rectangles, "name", 
           link_islands_k = 1) |> 
  st_manual_join_nb(3,4) |> 
  st_quickmap_nb(nodes = "numeric")
\end{verbatim}

@@

\begin{verbatim}
# remove an existing connection using unit name, not index

st_bridges(rectangles, "name", 
           link_islands_k = 1) |> 
  st_manual_join_nb(3,4) |> 
  st_manual_cut_nb("Rect1","Rect2") |> 
  st_quickmap_nb(nodes = "numeric")
\end{verbatim}

@@

The use of these neighbourhood structures is particularly associated with CAR-type models.
These are often implemented in a Bayesian framework using
\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. For example, the \CRANpkg{brms} ICAR structure requires the neighbourhood relationships to be in matrix form. The pre-functions will output the
neighbourhood structure in the desired format for use in any of these
frameworks. A convenient frequentist alternative is to use the
\CRANpkg{mgcv} package which requires a named list of neighbours. It has the functionality to create such models
using \texttt{bs="mrf"}. It also has the ability to combine these with a
hierarchical structure using \texttt{bs="re"}. While the outputs from the
Bayesian structures mentioned above can be extracted in the same way as
any other component of the model, it can be somewhat awkward to get the
estimates from \CRANpkg{mgcv} model. \texttt{sfislands} has two post-functions to conveniently extract and visualise these.

@@

\caption{\label{tab:postfunc-latex}Post-functions: tidy estimates from mgcv.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l|l}
\hline
\textbf{function} & \textbf{purpose}\\
\hline
st\_augment() & augment the original dataframe with model predictions\\
\hline
st\_quickmap\_preds() & generate quick maps of these predictions\\
\hline
\end{tabular}
\end{table}

@@

This function augments the original dataframe with the estimated means
and standard errors of the spatially varying predictions from a fitted \CRANpkg{mgcv} model in a similar manner to
how the \CRANpkg{broom} package (Robinson, Hayes, and Couch 2023) operates. The \texttt{geometry} column, as per convention, remains as the last column of the augmented dataframe, while the predictions are positioned immediately before it. \footnote{In a similar way, \texttt{st\_augment()} can also be used to append the random effects from \texttt{lme4} and \texttt{nlme} models to an \texttt{sf} dataframe, which can then be easily mapped using \texttt{st\_quickmap\_preds()}.} The spatially varying predictions which \texttt{st\_augment()} extracts from an \CRANpkg{mgcv} model are

\begin{itemize}
\tightlist
\item
  random effects (which are called in \CRANpkg{mgcv} with \texttt{bs=\textquotesingle{}re\textquotesingle{}}), and
\item
  ICAR components (\texttt{bs=\textquotesingle{}mrf\textquotesingle{}}).
\end{itemize}

Consider the model structure described in the code below using \CRANpkg{mgcv} syntax. In this model \(y\) is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a set of random intercepts and slopes for the covariate at a \emph{region} level, and a set of ICAR varying intercepts and slopes at a lower \emph{sub-region} level.

\begin{verbatim}
# creating an mgcv model

mgcv::gam(
  y ~ covariate +                   # fixed intercept and effect for covariate  
    s(region, bs = "re") +                  # random intercept at level region
    s(region, covariate, bs = "re") +          # random slopes at level region
    s(sub-region, 
      bs = 'mrf',
      xt = list(nb = data$nb), 
      k = k) +                    # ICAR varying intercept at level sub-region
    s(sub-region, by = covariate, 
      bs = 'mrf',
      xt = list(nb = data$nb), 
      k = k),           # ICAR varying slope for covariate at level sub-region
  data = data, 
  method = "REML")
\end{verbatim}

When labelling the new prediction columns which are augmented to the original dataframe from such a model, \texttt{st\_augment()} follows the formula syntax of the \CRANpkg{lme4} package (Bates et al. 2015), where the pipe symbol (\texttt{\textbar{}}) indicates ``\emph{grouped by}''. Table \ref{tab:staugtab-latex} shows how the augmented columns in this scenario would be named. Each column name begins with either \texttt{random.effect.} or \texttt{mrf.smooth.} as appropriate. An additional column is also added for the standard error of each estimate, as calculated by \CRANpkg{mgcv}. These columns are named as above but with \texttt{se.} prepended (e.g.~\texttt{se.random.effect.region}).

@@

\caption{\label{tab:staugtab-latex}The naming procedure for augmented columns from different mgcv structures.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l|l}
\hline
\textbf{mgcv syntax} & \textbf{column name}\\
\hline
s(region, bs = 're') & random.effect.region\\
\hline
s(region, covariate, bs = 're') & random.effect.covariate|region\\
\hline
s(sub-region, bs = 'mrf', xt = list(nb = data\$nb) & mrf.smooth.sub-region\\
\hline
s(sub-region, by = covariate, bs = 'mrf', xt = list(nb = data\$nb) & mrf.smooth.covariate|sub-region\\
\hline
\end{tabular}
\end{table}

@@

These estimates can then be quickly mapped. As it is possible to include more than 1 spatially varying component, the output of this function is a list of plots. They can be viewed individually by indexing, or all at one using, for example, the \texttt{plotlist} argument from the \texttt{ggarrange()} function which is part of the \CRANpkg{ggpubr} (Kassambara 2023) package. We will see this function in practice in the following examples. The maps which it generates are automatically titled and subtitled according to the type of effect. For example, the map showing predictions for \texttt{random.effect.region} will have ``\emph{region}'' as its title and ``\emph{random.effect}'' as its subtitle.

@@

The data for this section have been downloaded from the National Earthquake Information Center, \href{https://earthquake.usgs.gov/earthquakes/search/}{USGS earthquake
catalogue}. The
datasets with accompanying explanations are available at
\url{https://github.com/horankev/quake_data}. They capture all recorded earthquakes in
and close to Indonesia from the beginning of January 1985 to the end of December 2023. Figure \ref{fig:fault-buffers} shows a map of Indonesia, divided into 33 provinces, with other neighbouring or bordering countries filled in grey. The many local
faults which lie within 300km of the shore are shown in yellow with green outlines.

@@

To get an interpretable measure of the concentration of faults in any area,
these faults are transformed from linestrings to polygons by setting a buffer of
10km around them, which explains their green outline. Now both our faults and the sizes of provinces are in units of kilometres
squared. This means we can generate a unitless metric of what proportion of any administrative unit
is covered by these buffered faults. This measure across provinces is shown in Figure
\ref{fig:fault-conc}.

@@

Earthquake incidence per province has been calculated as the total number of earthquakes with
an epicentre within that province per unit area. We have
restricted counts to earthquakes \textgreater5.5 on the moment magnitude scale, which is the point
at which they are often labelled as potentially damaging.

The occurrences of these earthquakes are shown in Figure \ref{fig:quake-occur}, their total per province in Figure \ref{fig:quake-totals}, and finally, their incidence or count per square kilometre can be seen in Figure \ref{fig:quake-conc}.

@@

\caption{Earthquakes in Indonesia of magnitude > 5.5, 1985-2023. Categorised by magnitude as medium, large or extra-large. }\label{fig:quake-occur}
\end{figure}

@@

As this is count data, we will model it as a poisson distribution with
\(\lambda\) as the mean count per province. For \(i = 1,...,n\) provinces,
the dependent variable in this model is
\begin{equation}
y_i = \text{earthquake count}_i 
\label{eq:eq1}
\end{equation}
while the explanatory variable is
\begin{equation}
x_i = \text{fault concentration}_i = \frac{\text{area of buffered faults in province}_i} {\text{province area}_i}.
\label{eq:eq2}
\end{equation}
Firstly, when excluding the incidence and just modelling counts, where
\(y_i = \text{earthquake count in province}_i\), the poisson model is of
the following form:
\begin{equation}
y_i | \lambda_i \sim \text{Pois}(\lambda_i)
\label{eq:eq3}
\end{equation}
with
\begin{equation}
E(y_i | \lambda_i) = \lambda_i.
\label{eq:eq4}
\end{equation}
We model
\begin{equation}
log(\lambda_i) = \beta_0 + \beta_1x_i + \gamma_i.
\label{eq:eq5}
\end{equation}
where \(\gamma_i\)
is a term with a correlation structure reflecting a province's location
relative to other provinces.

We can describe these relationships by setting up a neighbourhood
structure based on queen contiguity where a pair of provinces are
considered neighbours if they share at least one point of boundary. This can be modelled as a Markov random field to generate an ICAR model with a spatially varying term. Each of these terms will be correlated with the others according to the neighbourhood structure we have defined.

@@

We now include an offset term (here, area) because we are more interested
in modelling the incidence than in the actual count, such that
\begin{equation}
log(\frac{\lambda_i} {\text{area}_i}) = \beta_0 + \beta_1x_i + \gamma_i
\label{eq:eq6}
\end{equation}
which is equivalent to
\begin{equation}
log(\lambda_i) = \beta_0 + \beta_1x_i + \gamma_i + log(\text{area}_i).
\label{eq:eq7}
\end{equation}
We are still modelling \(log(\lambda)\) rather than the incidence, but we
are adding an offset to adjust for differing areas. Modelling \(log(\lambda)\) and adding an offset is equivalent to
modelling incidence, and coefficients can be interpreted that way.

@@

Having described the type of model we wish to implement, we now show how \texttt{sfislands} can be used to streamline the process.

@@

Having made some manual adjustments to the neighbourhood structure, the
new structure can be seen in Figure \ref{fig:ind-contig3}. Edge effects
have also been mitigated by imposing additional connections on the two extreme
provinces (1 and 23), which would otherwise have only one neighbour, so that they now also include
their two next closest neighbours.

@@

We now create the ICAR model using, in this case, the \CRANpkg{mgcv} package. We will be able to use the output of \texttt{st\_bridges}, which we have named \texttt{prep\_data}, as both the data source for the model and the neighbourhood structure (by specifying the column \texttt{nb} which contains the neighbourhood list).

@@

\hypertarget{workflow-summary}{%
\subsection{Workflow summary}\label{workflow-summary}}
